---
title: "2022-Franklin-Trawl"
output: html_document
date: "2022-07-26"
---

```{r setup, include=FALSE}
library(tidyverse)
library(lubridate)
library(readxl)
library(here)
library(lubridate)
library(obistools)
```

Download file:

```{r download, eval = FALSE}
download.file("https://github.com/international-year-of-the-salmon/2022-Franklin-Data-Template/blob/main/IYS_2022_FRANKLIN.xlsx?raw=true", here::here("IYS_data_template", "IYS_2022_FRANKLIN.xlsx"), quiet = TRUE, mode = "wb", overwrite = TRUE)

Franklin_event <- read_excel(here("IYS_data_template", "IYS_2022_FRANKLIN.xlsx"), sheet = "3. SAMPLING EVENT INFO") %>% filter(Event_Type == "Trawl")
```

Create function to determine polygon coordinates required for the metadata:

```{r polygon coordinates, include = FALSE}
polygon_coords <- function(Franklin_event){
  Franklin_event <- Franklin_event %>% select(Latitude_Start_DecDeg, Longitude_Start_DecDeg) %>% 
    drop_na(Latitude_Start_DecDeg, Longitude_Start_DecDeg) 
  ch <- chull(Franklin_event)
  coords <- Franklin_event[c(ch, ch[1]), ]
  coords <- paste(coords$Latitude_Start_DecDeg, coords$Longitude_Start_DecDeg, sep = ",", collapse = " ")
  coords
}
polygon_coords(Franklin_event)
```

Create Event Core:

```{r event_core, include = FALSE}
Franklin_event <- Franklin_event %>% 
  mutate(eventDate = paste(Time_Start_UTC, Time_End_UTC, sep = "/"),
         Month = ifelse(Month == "Feb", "2", "3") %>% as.numeric()) # Change Feb to 2, and Mar to 3.

# Change Station_Event_IDs to facilitate project-level integration:
Franklin_event <- Franklin_event %>%
  dplyr::rename(eventID = DFO_Station_Event_ID)
    
Franklin_event_coordinates <- Franklin_event %>%
  select(eventID,
         Longitude_Start_DecDeg, Latitude_Start_DecDeg,
         Longitude_End_DecDeg, Latitude_End_DecDeg) %>%
  distinct(eventID, .keep_all = TRUE) %>%
  mutate(footprintWKT = paste("LINESTRING (", Longitude_Start_DecDeg, Latitude_Start_DecDeg, ",", 
                              Longitude_End_DecDeg, Latitude_End_DecDeg, ")")) 

Franklin_event_linestring <- obistools::calculate_centroid(Franklin_event_coordinates$footprintWKT)
Franklin_event_linestring <- cbind(Franklin_event_coordinates, Franklin_event_linestring) %>%
  select(eventID, footprintWKT, decimalLatitude, decimalLongitude, coordinateUncertaintyInMeters)

Franklin_event_trawl <- Franklin_event %>%
  select(eventID,
         eventDate,
         Year, Month, Day,
         minimumDepthInMeters = Minimum_Sampling_Depth_Meters,
         maximumDepthInMeters = Maximum_Sampling_Depth_Meters,
         sampleSizeValue = Sampling_Duration,
         sampleSizeUnit = Sampling_Duration_units) %>%
  left_join(Franklin_event_linestring, by = "eventID") %>%
  distinct(eventID, .keep_all = TRUE) %>%
  mutate(geodeticDatum = "WGS84",
         samplingEffort = "midwater trawl",
         institutionCode = "PBS",
         institutionID = "https://edmo.seadatanet.org/report/4180",
         datasetID = "https://doi.org/10.21966/gmrz-ad56",
         bibliographicCitation = "King, J. R., Tabata, A., Freshwater, C., & Zubkowski, T. (2022). Trawl Data from the CCGS Sir John Franklin during the 2022 International Year of the Salmon Pan-Pacific Winter High Seas Expedition [Dataset]. North Pacific Anadromous Fish Commission. https://doi.org/10.21966/gmrz-ad56")

# Save locally:
write_csv(Franklin_event_trawl, here("standardized_data", "Franklin2022_event.csv"))

# Make sure there are no duplicated EventIDs:
Franklin_event_trawl[duplicated(Franklin_event_trawl$eventID),] # should be none. 
```

Next, create the occurrence extension for the overall catch: 

```{r occurrence overall catch, eval = FALSE}
occurrence_allCatch <- read_excel(here("IYS_data_template", "IYS_2022_FRANKLIN.xlsx"), sheet = "4. CATCH_FINAL INFO")

occurrence_allCatch$verbatimIdentification <- occurrence_allCatch$Scientific_Name
occurrence_allCatch <- occurrence_allCatch %>% rename(scientificname = Scientific_Name)
unique_spp <- unique(occurrence_allCatch$scientificname) %>% as.data.frame()
colnames(unique_spp) <- "scientificname"

# Assign WoRMS LSID to the unique species:
worms_id <- worrms::wm_records_names(unique(occurrence_allCatch$scientificname), marine_only = FALSE) %>% dplyr::bind_rows()

# Find out which species are not found in the WoRMS database:
worms_occ <- left_join(unique_spp, worms_id, by = "scientificname") %>% 
  filter(is.na(AphiaID)) %>% distinct(scientificname)

# Change the taxa names accordingly: 
occurrence_allCatch$scientificname <- gsub("Onychoteuthis borealijaponicus","Onychoteuthis borealijaponica", occurrence_allCatch$scientificname)
occurrence_allCatch$scientificname <- gsub("Unknown fish", "Pisces", occurrence_allCatch$scientificname)

# So now we can run worrms::wm_records_names again, and there should be no NA in the AphiaID column. Be sure to inspect the result as sometimes multiple AphiaIDs are generated for the same taxa.    
occurrence_allCatch_id <- worrms::wm_records_names(unique(occurrence_allCatch$scientificname), marine_only = FALSE) %>% bind_rows() 

# Omit certain AphiaIDs that do not apply:
occurrence_allCatch_id <- occurrence_allCatch_id %>% filter(!AphiaID %in% c("956054", "956065", "163921"))

# Now that all unique entries have a WoRMS ID, connect this to the original data frame:
Franklin_occurrence_all <- left_join(occurrence_allCatch_id, occurrence_allCatch, by = "scientificname")

# Omit biomass data from the occurrence extension:
Franklin_occurrence <- Franklin_occurrence_all %>%
  mutate(specificEpithet = stringr::word(scientificname, 2)) %>%
  select(eventID = DFO_Station_Event_ID, occurrenceID = DFO_Catch_ID, 
         scientificName = scientificname,
         verbatimIdentification,
         scientificNameID = `lsid`, 
         scientificNameAuthorship = authority,
         taxonomicStatus = status,
         taxonRank = rank, 
         vernacularName = Species_recorded, 
         individualCount = Catch_count, kingdom, phylum, class, order, family, genus, specificEpithet,
         dateIdentified = dateidentified, identifiedBy) %>% distinct() %>%
  mutate(basisOfRecord = "HumanObservation",
         occurrenceStatus = "present")

# Save locally:
write_csv(Franklin_occurrence, here("standardized_data", "IYS2022_Franklin_occurrence_trawl_all.csv"))
```

Next, create the occurrence extension for the specimen catch: 

```{r occurrence overall catch, eval = FALSE}
occurrence_specimen <- read_excel(here("IYS_data_template", "IYS_2022_FRANKLIN.xlsx"), sheet = "5. SPECIMEN INFO")

occurrence_specimen$verbatimIdentification <- occurrence_specimen$Scientific_Name
occurrence_specimen <- occurrence_specimen %>% rename(scientificname = Scientific_Name)
unique_spp_specimen <- unique(occurrence_specimen$scientificname) %>% as.data.frame()
colnames(unique_spp_specimen) <- "scientificname"

# Assign WoRMS LSID to the unique species:
worms_id <- worrms::wm_records_names(unique(occurrence_specimen$scientificname), marine_only = FALSE) %>% dplyr::bind_rows()

# Find out which species are not found in the WoRMS database:
worms_occ <- left_join(unique_spp_specimen, worms_id, by = "scientificname") %>% 
  filter(is.na(AphiaID)) %>% distinct(scientificname)

# Change the taxa names accordingly: 
occurrence_allCatch$scientificname <- gsub("Onychoteuthis borealijaponicus","Onychoteuthis borealijaponica", occurrence_allCatch$scientificname)
occurrence_allCatch$scientificname <- gsub("Unknown fish", "Pisces", occurrence_allCatch$scientificname)

# So now we can run worrms::wm_records_names again, and there should be no NA in the AphiaID column. Be sure to inspect the result as sometimes multiple AphiaIDs are generated for the same taxa.    
occurrence_specimen_id <- worrms::wm_records_names(unique(occurrence_specimen$scientificname), marine_only = FALSE) %>% bind_rows() 

# Omit certain AphiaIDs that do not apply:
occurrence_specimen_id <- occurrence_specimen_id %>% filter(!AphiaID %in% c("956054", "956065", "163921"))

# Now that all unique entries have a WoRMS ID, connect this to the original data frame:
Franklin_occurrence_specimen <- left_join(occurrence_specimen_id, occurrence_specimen, by = "scientificname")

# Omit biomass data from the occurrence extension:
Franklin_occ_specimen <- Franklin_occurrence_specimen %>%
  mutate(specificEpithet = stringr::word(scientificname, 2)) %>%
  select(eventID = DFO_Station_Event_ID, 
         occurrenceID = DFO_Specimen_ID, 
         scientificName = scientificname,
         verbatimIdentification,
         scientificNameID = `lsid`, 
         scientificNameAuthorship = authority,
         taxonomicStatus = status,
         taxonRank = rank, 
         lifeStage = Ageclass,
         vernacularName = common_name, 
         kingdom, phylum, class, order, family, genus, specificEpithet) %>% distinct() %>%
  mutate(basisOfRecord = "HumanObservation",
         occurrenceStatus = "present")

# Save locally:
write_csv(Franklin_occ_specimen, here("standardized_data", "IYS2022_Franklin_occurrence_trawl_specimen.csv"))
```

Combine both occurrence extensions into a final data table:

```{r occurrence, eval = FALSE}
Franklin2022_occ <- dplyr::bind_rows(Franklin_occurrence, Franklin_occ_specimen)

# To re-order the occurrenceID, use following code:
order <- stringr::str_sort(Franklin2022_occ$occurrenceID, numeric=TRUE)
Franklin2022_occ <- Franklin2022_occ[match(order, Franklin2022_occ$occurrenceID),]

# Remove NA and replace with empty cells:
Franklin2022_occ <- sapply(Franklin2022_occ, as.character)
Franklin2022_occ[is.na(Franklin2022_occ)] <- ""
Franklin2022_occ <- as.data.frame(Franklin2022_occ)

# Check if occurrenceIDs are all unique - answer should be TRUE:
length(unique(Franklin2022_occ$occurrenceID)) == nrow(Franklin2022_occ)
Franklin2022_occ[duplicated(Franklin2022_occ$occurrenceID),]

# Save the occurrence Core locally and in Google Drive:
write_csv(Franklin2022_occ, here("standardized_data", "Franklin2022_occ.csv"))
```

Create resourceRelationship extension: 

```{r resourceRelationship, eval = FALSE}
Franklin2022_resourceRelationship <- Franklin2022_occ %>%
  select(eventID, occurrenceID, scientificName, individualCount)
Franklin2022_resourceRelationship$individualCount <- as.numeric(Franklin2022_resourceRelationship$individualCount)
  
Franklin2022_resourceRelationship <- Franklin2022_resourceRelationship %>%
  mutate(resourceID = ifelse(is.na(individualCount), Franklin2022_resourceRelationship$occurrenceID, NA)) %>%
  group_by(eventID, scientificName) %>%
  filter(n() != 1) %>%
  ungroup()

Franklin2022_resourceRelationship <- Franklin2022_resourceRelationship %>%
  mutate(relatedResourceID = ifelse(is.na(individualCount), NA, Franklin2022_resourceRelationship$occurrenceID)) %>%
  mutate(relationshipOfResource = ifelse(!is.na(resourceID), "is a subset of", NA)) %>%
  dplyr::arrange(eventID, scientificName) %>%
  fill(relatedResourceID) %>%
  filter(!is.na(resourceID))

order <- stringr::str_sort(Franklin2022_resourceRelationship$resourceID, numeric = TRUE)
Franklin2022_resourceRelationship <- Franklin2022_resourceRelationship[match(order, Franklin2022_resourceRelationship$resourceID),]
Franklin2022_resourceRelationship <- Franklin2022_resourceRelationship %>%
  mutate(resourceRelationshipID = paste(relatedResourceID, "rr", sep = "-"),
         ID = sprintf("%03d", row_number()),
         resourceRelationshipID = paste(resourceRelationshipID, ID, sep = "-")) %>%
  select(eventID, resourceRelationshipID, resourceID, relationshipOfResource, relatedResourceID)

write_csv(Franklin2022_resourceRelationship, here("standardized_data", "Franklin2022_resourceRelationship.csv"))
```

Create eMOF extension data table:

```{r eMOF, eval = FALSE}
# Data table for the facts related to the Research Vessel:
Franklin2022_rv <- Franklin_event %>%
  select(eventID = Station_Event_ID) %>%
  mutate_all(as.character) %>%
  mutate(measurementID = paste(eventID, "rv", sep = "-"), 
         measurementType = "Research Vessel",
         measurementTypeID = "http://vocab.nerc.ac.uk/collection/L06/current/31/",
         measurementValue = "CCGS Sir John Franklin",
         measurementValueID = "http://vocab.nerc.ac.uk/collection/C17/current/18KD/",
         measurementUnit = NA,
         measurementUnitID = NA) %>% distinct()

# Data table for the bottom depth at each sampling station:
Franklin2022_bottomdepth <- Franklin_event %>%
  select(eventID = Station_Event_ID,
         Bottom_Depth_Meters = Bottom_Depth_Start) %>%
  mutate_all(as.character) %>%
  pivot_longer(cols = Bottom_Depth_Meters,
               names_to = "measurementType",
               values_to = "measurementValue") %>%
  mutate(measurementTypeID = case_when(
           measurementType == "Bottom_Depth_Meters" ~ "http://vocab.nerc.ac.uk/collection/P09/current/BATH/"),
         measurementUnit = "meters",
         measurementUnitID = case_when(
           measurementUnit == "meters" ~ "http://vocab.nerc.ac.uk/collection/P06/current/ULAA/"),
         measurementValueID = NA,
         measurementID = paste(eventID, measurementType, sep = "-")) %>%
  select(eventID, measurementID, measurementType, measurementTypeID, measurementValue, measurementValueID,
         measurementUnit, measurementUnitID)

# Data table for the sampling effort:
Franklin2022_samplingEffort <- Franklin_event %>%
  select(eventID = Station_Event_ID,
         Sampling_Duration,
         Tow_speed,
         Tow_distance) %>%
  mutate_all(as.character) %>%
  pivot_longer(cols = Sampling_Duration:Tow_distance,
               names_to = "measurementType",
               values_to = "measurementValue") %>%
  mutate(measurementTypeID = case_when(
           measurementType == "Sampling_Duration" ~ "http://vocab.nerc.ac.uk/collection/P01/current/AZDRZZ01/",
           measurementType == "Tow_speed" ~ "http://vocab.nerc.ac.uk/collection/P01/current/TOWSPEED/",
           measurementType == "Tow_distance" ~ "http://vocab.nerc.ac.uk/collection/P01/current/LENTRACK/"),
         measurementUnit = case_when(
           measurementType == "Sampling_Duration" ~ "minutes",
           measurementType == "Tow_speed" ~ "kilometers per hour",
           measurementType == "Tow_distance" ~ "kilometers"),
         measurementUnitID = case_when(
           measurementUnit == "minutes" ~ "http://vocab.nerc.ac.uk/collection/P06/current/UMIN/",
           measurementUnit == "kilometers per hour" ~ "https://vocab.nerc.ac.uk/collection/P06/current/KMHR/",
           measurementUnit == "kilometers" ~ "http://vocab.nerc.ac.uk/collection/P06/current/ULKM/"),
         measurementValueID = NA,
         measurementID = paste(eventID, measurementType, sep = "-")) %>%
  select(eventID, measurementID, measurementType, measurementTypeID, measurementValue, measurementValueID,
         measurementUnit, measurementUnitID)

# Data table for the oceanic/atmospheric conditions during the sampling event:
Franklin2022_conditions <- Franklin_event %>%
  select(eventID = Station_Event_ID,
         Wave_height) %>%
  mutate_all(as.character) %>%
  pivot_longer(cols = Wave_height,
               names_to = "measurementType",
               values_to = "measurementValue") %>%
  mutate(measurementTypeID = case_when(
           measurementType == "Wave_height" ~ "http://vocab.nerc.ac.uk/collection/P07/current/JNQS0CMX/"),
         measurementUnit = case_when(
           measurementType == "Wave_height" ~ "meters"),
         measurementUnitID = case_when(
           measurementUnit == "meters" ~ "http://vocab.nerc.ac.uk/collection/P06/current/ULAA/"),
         measurementValueID = NA,
         measurementID = paste(eventID, measurementType, sep = "-")) %>% 
  drop_na(measurementValue) %>%
  select(eventID, measurementID, measurementType, measurementTypeID, measurementValue, measurementValueID,
         measurementUnit, measurementUnitID)

Franklin2022_trawl_allCatch_eMOF <- Franklin_occurrence_all %>%
  select(eventID = Station_Event_ID,
         occurrenceID = Catch_ID,
         individualCount = Catch_count,
         Catch_weight,
         Count_method) %>%
  mutate_all(as.character) %>%
  pivot_longer(cols = c(individualCount:Catch_weight),
               names_to = "measurementType",
               values_to = "measurementValue") %>%
  rename(measurementMethod = Count_method)

Franklin2022_trawl_allCatch_eMOF <- Franklin2022_trawl_allCatch_eMOF %>%
  mutate(measurementID = paste0(occurrenceID, measurementType, sep = "-"),
         measurementTypeID = case_when(
           measurementType == "individualCount" ~ "http://vocab.nerc.ac.uk/collection/P01/current/OCOUNT01/",
           measurementType == "Catch_weight" ~ "http://vocab.nerc.ac.uk/collection/S06/current/S0600088/"),
         measurementUnit = case_when(
           measurementType == "individualCount" ~ "individuals",
           measurementType == "Catch_Weight" ~ "kilogram"),
         measurementUnitID = case_when(
           measurementUnit == "kilogram" ~ "http://vocab.nerc.ac.uk/collection/P06/current/KGXX/",
           measurementUnit == "individuals" ~ "http://vocab.nerc.ac.uk/collection/P06/current/UUUU/"),
         measurementValueID = NA) %>%
  select(eventID, occurrenceID, measurementID, measurementType, measurementTypeID, measurementValue, measurementValueID,
         measurementUnit, measurementUnitID, measurementMethod)

# And for the individual specimen data:
Franklin2022_ind_eMOF <- Franklin_occurrence_specimen %>%
  select(eventID = Station_Event_ID,
         occurrenceID = Specimen_ID, 
         Ageclass, Specimen_Length, Specimen_Weight, Sex, Length_Type) %>%
  mutate_all(as.character) %>%
  pivot_longer(cols = Ageclass:Sex,
               names_to = "measurementType", 
               values_to = "measurementValue") %>%
  mutate(measurementID = paste(occurrenceID, measurementType, sep = "-"),
         measurementType = ifelse(measurementType == "Specimen_Length" & Length_Type == "Total Length", "Specimen total length", measurementType),
         measurementType = ifelse(measurementType == "Specimen_Length" & Length_Type == "Fork Length", "Specimen fork length", measurementType),
         measurementType = ifelse(measurementType == "Specimen_Length" & Length_Type == "Mantle Length", "Specimen mantle length", measurementType),
         measurementType = ifelse(measurementType == "Specimen_Length" & Length_Type == "Bell Diameter", "Specimen bell diameter", measurementType),
         measurementType = ifelse(measurementType == "Specimen_Length" & Length_Type == "Standard Length", "Specimen standard length", measurementType)) %>%
  mutate(measurementTypeID = case_when(
           measurementType == "Ageclass" ~ "http://vocab.nerc.ac.uk/collection/P01/current/LSTAGE01/",
           measurementType == "Sex" ~ "http://vocab.nerc.ac.uk/collection/P01/current/ENTSEX01/",
           measurementType == "Specimen total length" ~ "http://vocab.nerc.ac.uk/collection/P01/current/TL01XX01",
           measurementType == "Specimen fork length" ~ "http://vocab.nerc.ac.uk/collection/P01/current/FL01XX01/",
           measurementType == "Specimen mantle length" ~ "http://vocab.nerc.ac.uk/collection/P01/current/DML1XX01/",
           measurementType == "Specimen bell diameter" ~ "http://vocab.nerc.ac.uk/collection/P01/current/OBSINDDM/",
           measurementType == "Specimen standard length" ~ "http://vocab.nerc.ac.uk/collection/P01/current/SL01XX01/",
           measurementType == "Specimen_Weight" ~ "http://vocab.nerc.ac.uk/collection/S06/current/S0600088/"),
         measurementUnit = case_when(
           measurementType %in% c("Specimen total length", "Specimen fork length", "Specimen mantle length", 
                                  "Specimen bell diameter", "Specimen standard length") ~ "millimeters",
           measurementType == "Specimen_Weight" ~ "grams"),
         measurementUnitID = case_when(
           measurementUnit == "grams" ~ "http://vocab.nerc.ac.uk/collection/P06/current/UGRM/",
           measurementUnit == "millimeters" ~ "http://vocab.nerc.ac.uk/collection/P06/current/UXMM/"),
         measurementValueID = case_when(
           measurementValue == "Unknown" ~ "https://vocab.nerc.ac.uk/collection/S10/current/S105/",
           measurementValue == "Immature" ~ "http://vocab.nerc.ac.uk/collection/S11/current/S1171/",
           measurementValue == "Male" ~ "https://vocab.nerc.ac.uk/collection/S10/current/S103/",
           measurementValue == "Female" ~ "http://vocab.nerc.ac.uk/collection/S10/current/S102/")) %>%
  select(eventID, occurrenceID, measurementID, measurementType, measurementTypeID, measurementValue, measurementValueID,
         measurementUnit, measurementUnitID)
```

Combine the extended measurementOrFact extensions into a single eMOF extension, and save locally and in GoogleDrive:

``` {r}
Franklin2022_eMOF <- dplyr::bind_rows(Franklin2022_rv,
                                     Franklin2022_bottomdepth,
                                     Franklin2022_samplingEffort,
                                     Franklin2022_conditions,
                                     Franklin2022_trawl_allCatch_eMOF,
                                     Franklin2022_ind_eMOF)

# Remove NA and replace with empty cells:
Franklin2022_eMOF <- sapply(Franklin2022_eMOF, as.character)
Franklin2022_eMOF[is.na(Franklin2022_eMOF)] <- ""
Franklin2022_eMOF <- as.data.frame(Franklin2022_eMOF)

# Save locally
write_csv(Franklin2022_eMOF, here("standardized_data", "Franklin2022_eMOF.csv"))
```